---
title: "Reproducible Research - Project 1"
author: "NML DesJardins"
date: "August 11, 2015"
output:
  html_document:
    keep_md: yes
---

# Loading and preprocessing the data  

**Load the data:**  

```{r}
setwd("~/Downloads")
data<-read.csv("activity.csv", na.strings="NA")
head(data)
```

**Reformat date from factor to date:**  

```{r}
data$date<-as.Date(data$date)
str(data)
```

# What is mean total number of steps taken per day?

**Compute the total number of steps per day:**  

```{r, message=F, warning=F}
library(dplyr)
steps.per.date<-data%>%
        group_by(date)%>%
        do({
               x<-as.data.frame(sum(.$steps,na.rm=T))
        })
colnames(steps.per.date)<-c("date", "total_steps")
steps.per.date
```

**Make a historgram of the total number of steps taken each day:**  

```{r}
hist(steps.per.date$total_steps, main="Number of Steps per Day", xlab="Steps", 
     col="purple")
```

```{r}
mean.steps<-round(mean(steps.per.date$total_steps, na.rm=T))
median.steps<-median(steps.per.date$total_steps,na.rm=T)
hilo<-ifelse(mean.steps>median.steps, "lower","higher")
```

**On average, this person took `r mean.steps` steps per day. The median number of 
steps was somewhat `r hilo`, at `r median.steps` steps per day.**  

******

# What is the average daily activity pattern?  

**Compute average number of steps within each interval:**  

```{r}
steps.per.interval<-data%>%
        group_by(interval)%>%
        do({
               x<-as.data.frame(mean(.$steps,na.rm=T))
        })
colnames(steps.per.interval)<-c("interval", "avg_steps")
steps.per.interval
```

**Time series plot of average steps per interval, across all days:**  

```{r}
with(steps.per.interval, plot(interval, avg_steps, type="l", xlab="Interval", 
                              ylab="Average Number of Steps", 
                              main="Number of Steps by Time of Day", col="purple"))
```

```{r}
maxstep<-steps.per.interval[which(steps.per.interval$avg_steps==
                                          max(steps.per.interval$avg_steps)),]
```

**On average, this person took the most steps during the interval at `r maxstep$interval`.**    
  
******

# Imputing Missing Values   

```{r}
nmiss<-table(is.na(data))[2]
```

**There are a total of `r nmiss` missing values in the dataset.**  
  
    
**Impute missing values with the mean for that interval to create a new dataset:** 
  
```{r}
data.mi<-merge(data,steps.per.interval)
data.mi$steps=with(data.mi,ifelse(is.na(steps==T),avg_steps,steps))
table(is.na(data.mi))
head(data.mi)
```

**Compute total number of steps each day:**  
```{r}
steps.per.date.mi<-data.mi%>%
        group_by(date)%>%
        do({
               x<-as.data.frame(sum(.$steps))
        })
colnames(steps.per.date.mi)<-c("date", "total_steps_mi")
steps.per.date.mi
```


**Histogram of the total number of steps each day:**  
```{r}
with(steps.per.date.mi,hist(total_steps_mi, main="Number of Steps per Day", 
                            xlab="Steps (mean imputed)", col="blue"))
                            
```
```{r}
options(scipen=999)
mean.steps.mi<-round(mean(steps.per.date.mi$total_steps_mi))
median.steps.mi<-round(median(steps.per.date.mi$total_steps_mi))

hilo.mi<-ifelse(mean.steps.mi>median.steps.mi,"higher than",
                ifelse(mean.steps.mi<median.steps.mi,"lower than","equal to"))

skew<-ifelse(hilo.mi=="equal to", "The mean and median are now equivalent, 
             indicating that the data are no longer skewed.","As in the raw data, 
             the mean and median are not equivalent, indicating that the data 
             are still at least somewhat skewed.")

hilo_mean<-ifelse(mean.steps.mi>mean.steps, "higher than",
                  ifelse(mean.steps.mi<mean.steps,"lower than", "equal to"))
hilo_mdn<-ifelse(median.steps.mi>median.steps, "higher than",
                 ifelse(median.steps.mi<median.steps,"lower than","equal to"))

```
**After replacing missing values with the means for each interval, this person took, on average, `r mean.steps.mi` steps per day, which is `r hilo_mean` the mean obtained from the raw data. The median number of steps (after mean imputation) was `r median.steps.mi`, which is `r hilo_mdn` the median obtained from the raw data. `r skew`**  
  
    
******

# Differences in activity patterns between weekends and weekdays  
  
**Create a factor variable for weekdays vs. weekends:**  
```{r}
data.mi$daytype<-as.factor(with(data.mi, ifelse(weekdays(data$date)=="Sunday"| 
                                              weekdays(data$date)=="Saturday",
                                      "weekend","weekday")))
str(data.mi$daytype)
```


**Compute the mean number of steps for each interval across weekdays or weekends:**  
```{r}
steps.int.day<-data.mi%>%
        group_by(daytype,interval)%>%
        do({
               x<-as.data.frame(mean(.$steps))
        })
colnames(steps.int.day)<-c("daytype","interval", "avg_steps")
steps.int.day
```

**Create a time series plot of the average number of steps on weekdays and weekends:**  
```{r, message=F, warning=F}
library(ggplot2)
p<-ggplot(steps.int.day,aes(interval,avg_steps))
p + geom_line() + facet_grid(daytype~.) + labs(x = "Interval", y = "Number of steps")

